---
project:
  title: "ESS 330 Paper Proposals"
  output-dir: docs
  type: website
format:
  
  html:
    self-contained: true
authors:
  - name: Mia Colangelo
    affiliation: Colorado State University
    roles: writing
    corresponding: true
  - name: Neva Morgan
    affiliation: Colorado State University
    roles: writing
    corresponding: false
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

## **Title:**

### **Flight in Flux:** Analyzing American Bittern (Botaurus lentiginosus) Migration Timing and Climate Change

#### **Introduction:**

Climate change poses many questions relating to systems across the
globe, especially ecological and evolutionary systems. One system which
could potentially be heavily impacted is migration. Bird migration
timing is critical for ensuring species survival and ecological balance.
As anthropogenic climate change has altered seasonal patterns, there is
a potential disruption to migration timing, impacting birds' breeding
success, food availability, and overall population health. The American
Bittern (Botaurus lentiginosus), for example, depends on the thriving
vegetation of marshlands for camouflage and nesting, as they are known
to nest on the ground. Wetland birds are expected to be affected
negatively by a considerable margin with the presence of a changing
climate altering their regular migratory ranges and patterns (Steen et
al., 2012).

#### **Abstract:**

The American Bittern is a solitary, cryptic species that relies on
wetlands for breeding and overwintering. Typically, these birds breed in
freshwater marshes across Canada and the northern United States before
migrating south to the southeastern U.S., Mexico, and Central America
for the winter. However, increasing global temperatures, shifts in
precipitation, and habitat loss due to climate change threaten to
disrupt these migration patterns. Past studies on migratory birds
indicate that climate change is leading to earlier spring arrivals,
delayed fall departures, and shifts in overall migration routes (La
Sorte & Thompson, 2007). For wetland-dependent birds like the American
Bittern, changes in water levels due to altered precipitation cycles
could further complicate their ability to find suitable stopover sites
and breeding grounds. Given that wetlands are already among the most
threatened ecosystems in North America, with significant losses due to
urbanization and agriculture, climate-induced changes pose an additional
challenge (Dahl, 2011). Studying the migration patterns of the American
Bittern under a changing climate is essential for predicting future
population trends and informing conservation strategies. By
understanding how climate variables influence their movements,
conservationists can prioritize habitat protection, restoration, and
adaptive management to mitigate potential negative impacts.

**Motivation:** Bird migration range and timing is critical for ensuring
species survival and ecological balance. However, climate change has
altered seasonal patterns, potentially disrupting migration timing and
impacting breeding success, food availability, and overall population
health. Understanding these shifts can inform conservation strategies
and mitigate risks to migratory bird species. Additionally, information
gained from analyzing one species can be used to make inferences about
the potential shifts in other similar species.

#### **Data**

Climate Data Online (NOAA Global Summary of the Month - U.S. Specific)

With NOAA’s Climate Data Online search, we focused on nine weather
stations located between Arizona (two weather stations: Phoenix and
Kingsman), California (three weather stations: Lake Tahoe, San Diego,
and Santa Rosa), Oregon (two weather stations: Chiloquin and Hermiston),
and Washington (two weather stations Shelton and Spokane), to map
multiple points of migratory status for the American Bittern. The files
were requested through eBird, and then downloaded into a readable CSV
for Excel to initially portion the data into the portions necessary for
our analysis. Each weather station collected data that contained the
specific information for the weather station, temperature,
precipitation, and wind-related metrics (of which are negligible for our
testing metrics).

Variables within the Datasets contain:

-   Metadata: Station name (NAME), location (LATITUDE, LONGITUDE,
    ELEVATION), and observation date (DATE)

-   Climate Variables: Includes average and extreme temperatures (e.g.,
    ADPT, ASLP, AWBT), precipitation statistics (e.g., DP01, DP10), and
    wind metrics (AWND)

-   Attributes and Flags: Columns such as \*\_ATTRIBUTES and logical
    flags provide metadata on data quality and source

-   Format: Character (station info and attribute flags), numeric
    (climate measures), and logical (indicator variables)

American Bittern eBird Data

This dataset was sourced from eBird and includes observations of
American Bitterns (Botaurus lentiginosus) reported by citizen
scientists. The dataset spans from 2000 to 2024 and includes over 70,000
observations.

-   Geographic Coverage: Arizona, California, Oregon, and Washington
    within the U.S.

-   Metadata: Observation Date Time Observations Started Latitude
    Longitude

-   Bird Details: Observation Count Breeding Code Behavior Code

-   Format: Primarily character and logical columns, with a few numeric
    fields (e.g., coordinates, observer count)



#### **Methods**

To examine the effects climate change will have on the migratory timing
of the American Bittern, we used a monthly summary of climate data from
nine recorded NOAA weather stations within Arizona, California, Oregon,
and Washington and observational data of their abundance within those
states from eBird.

Data Acquiring:

Climate data within nine weather stations stretched between Arizona,
California, Oregon, and Washington by NOAA Global Monthly Summaries,
were recorded on a scale between 1947-2025. By using ‘R’ resources, we
cleaned this data, removing unnecessary attributes for our study,
converting units to be consistent throughout each dataset, and removing
unnecessary data from past years to focus on data between 2000-2024.

Data Processing / Seasonal Aggregation:

Each weather station recorded the seasonal averages of temperature and
precipitation, which were then manipulated into a cleaned version of
that data, removing unnecessary wind records. With that, we created a
list item, using ‘RStudio’, that combined the contents of the
dataframes, to join them all together, ensuring columns and the contents
of the dataframes are not duplicated or N/Aed. We then integrated a
seasonal component into the data denoted by month: Winter = 12 (Dec.), 1
(Jan.), 2 (Feb.); Spring = 3 (Mar.), 4 (Apr.), 5 (May); Summer = 6
(Jun.), 7 (Jul.), 8 (Aug.); and Fall = 9 (Sep.), 10 (Oct.), 11 (Nov.).
The seasonal component was also applied to the American Bittern
migration dataframe, which was later converted to a tibble for later
analysis.

Migratory Timing Analysis of American Bittern:

Observational data was provided to understand the timing of Bitterns
within our four selected states, by grouping month and year to
understand temporal patterns of their appearance based on the seasonal
presence. Using a time series visualization to detect changes within the
migratory patterns based on their seasonality between 2000-2024.

Comparing Seasonal Trends of Climate Variability and Bittern
Observation:

To understand the correlation between seasonal changes for all the
weather stations, we averaged the temperature and precipitation of each
month per year. We then evaluated the long-term changes in bittern
observation from plot layouts used before, by visually plotting the
changes in the presence of climate seasonality.

#### Coding Component:
Beginning Mumbo Jumbo Libraries:

```{r}
library(readr) 
library(flextable) 
library(readr) 
library(tidyverse) 
library(tidymodels) 
library(ggplot2) 
library(purrr) 
library(dplyr) 
library(lubridate) 
library(tidyr) 
library(gridExtra) 
library(gapminder)
```

1.  Create time period for seasons

```{r}
# Read in all climate data
AZ_kingman <- read_csv("data/Climate Data/AZ-kingman-climate.csv")


AZ_phoenix <- read_csv("data/Climate Data/AZ-phoenix-climate.csv")


CA_lake_tahoe <- read_csv("data/Climate Data/CA-lake-tahoe-climate.csv")


CA_sandiego <- read_csv("data/Climate Data/CA-sandiego-climate.csv")


CA_santarosa <- read_csv("data/Climate Data/CA-santarosa_climate.csv")


OR_chiloquin <- read_csv("data/Climate Data/OR-chiloquin-climate.csv")


OR_hermiston <- read_csv("data/Climate Data/OR-hermiston-climate.csv")


WA_shelton <- read_csv("data/Climate Data/WA-shelton-climate.csv")


WA_spokane <- read_csv("data/Climate Data/WA-spokane-climate.csv")


# Clarify joining values:

AZ_kingman$LATITUDE <- as.numeric(AZ_kingman$LATITUDE)
AZ_kingman$LONGITUDE <- as.numeric(AZ_kingman$LONGITUDE)

AZ_phoenix$LATITUDE <- as.numeric(AZ_phoenix$LATITUDE)
AZ_phoenix$LONGITUDE <- as.numeric(AZ_phoenix$LONGITUDE)

CA_lake_tahoe$LATITUDE <- as.numeric(CA_lake_tahoe$LATITUDE)
CA_lake_tahoe$LONGITUDE <- as.numeric(CA_lake_tahoe$LONGITUDE)

CA_sandiego$LATITUDE <- as.numeric(CA_sandiego$LATITUDE)
CA_sandiego$LONGITUDE <- as.numeric(CA_sandiego$LONGITUDE)

CA_santarosa$LATITUDE <- as.numeric(CA_santarosa$LATITUDE)
CA_santarosa$LONGITUDE <- as.numeric(CA_santarosa$LONGITUDE)

OR_chiloquin$LATITUDE <- as.numeric(OR_chiloquin$LATITUDE)
OR_chiloquin$LONGITUDE <- as.numeric(OR_chiloquin$LONGITUDE)

OR_hermiston$LATITUDE <- as.numeric(OR_hermiston$LATITUDE)
OR_hermiston$LONGITUDE <- as.numeric(OR_hermiston$LONGITUDE)

WA_shelton$LATITUDE <- as.numeric(WA_shelton$LATITUDE)
WA_shelton$LONGITUDE <- as.numeric(WA_shelton$LONGITUDE)

WA_spokane$LATITUDE <- as.numeric(WA_spokane$LATITUDE)
WA_spokane$LONGITUDE <- as.numeric(WA_spokane$LONGITUDE)


#Cleaning Data - the SAGA:

AZ_kingman <- AZ_kingman %>%
  select(-contains("ATTRIBUTES"))

AZ_phoenix <- AZ_phoenix %>%
  select(-contains("ATTRIBUTES"))

CA_lake_tahoe <- CA_lake_tahoe %>%
  select(-contains("ATTRIBUTES"))

CA_sandiego <- CA_sandiego %>%
  select(-contains("ATTRIBUTES"))

CA_santarosa <- CA_santarosa %>%
  select(-contains("ATTRIBUTES"))

OR_chiloquin <- OR_chiloquin %>%
  select(-contains("ATTRIBUTES"))

OR_hermiston <- OR_hermiston %>%
  select(-contains("ATTRIBUTES"))

WA_shelton <- WA_shelton %>%
  select(-contains("ATTRIBUTES"))

WA_spokane <- WA_spokane %>%
  select(-contains("ATTRIBUTES"))

#Adding to a collective list to make join easier!
station_list <- list(AZ_kingman, AZ_phoenix, CA_lake_tahoe, CA_sandiego, CA_santarosa, OR_chiloquin, OR_hermiston, WA_shelton, WA_spokane)


cleaned_list <- map(station_list, ~ select(.x, -contains("ATTRIBUTES")))


#Binding the cleaned data!
climate_data <- bind_rows(cleaned_list)

```

a.  BUILDING SEASONS!:

```{r}

# Changing into month and year

head(climate_data$DATE)

climate_data <- climate_data %>%
  mutate(
    DATE = paste0(DATE, "-01"), # Adding a day component since it was missing!
    DATE = as.Date(DATE, format = "%Y-%m-%d"),
    year = year(DATE),
    month = month(DATE)
  )

# Creating a season:

climate_data <- climate_data %>%
  mutate(
    season = case_when(
      month %in% c(12, 1, 2)  ~ "Winter",
      month %in% c(3, 4, 5)   ~ "Spring",
      month %in% c(6, 7, 8)   ~ "Summer",
      month %in% c(9, 10, 11) ~ "Fall"
    ))



# Creating a season with the year to compare for climate
climate_data <- climate_data %>%
  mutate(
    season_year = if_else(month == 12, year + 1, year)
  )


#Separating temp and precp to clean further!

avg_temp <- climate_data %>%
  mutate(TAVG_F = (TAVG * 9/5) + 32) %>%
  group_by(STATION, NAME, month, season, season_year) %>%
  summarize(avg_temp = mean(TAVG_F, na.rm = TRUE), .groups = "drop") %>%
  filter(season_year >= 2000, season_year <= 2024)

avg_prec <- climate_data %>%
  group_by(STATION, NAME, month, season, season_year)%>%
  summarize(avg_prec = mean(PRCP, na.rm = TRUE), .groups = "drop") %>%
  filter(season_year >= 2000, season_year <= 2024)

```

2.  Understanding the changes of average temperature and precipitation
    in a map!


a.  Temperature Difference between 2000 - 2024 based on seasonality

```{r}
#TEMPERATURE!!!


#creating a separte coordination dataframe
station_coords <- climate_data %>%
  select(STATION, LATITUDE, LONGITUDE) %>%
  distinct()

#joining the two to relay to the map
avg_temp_map <- avg_temp %>%
  left_join(station_coords, by = "STATION")

#loading the u.s. map raster file!
us_map <- map_data("state")

#Creating map of avg temperature for Selected stations based on seasonality!

ggplot(avg_temp_map %>% filter(season_year >= 2000 & season_year <= 2024)) +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "white") +
  geom_point(aes(x = LONGITUDE, y = LATITUDE, color = avg_temp), size = 2) +
  scale_color_viridis_c(option = "plasma", name = "Avg Temp (°F)") +
  coord_fixed(1.3) +
  facet_wrap(~season) +
  labs(title = "Seasonal Average Temperatures by Station (2000–2024)",
       x = "", y = "") +
  theme_minimal()


# Let's compare first just what 2000 vs 2024 looks like!

# separating 2000 and 2024

avg_temp_2000 <- avg_temp_map %>% filter(season_year == 2000)
avg_temp_2024 <- avg_temp_map %>% filter(season_year == 2024)

# Create the ggplot for 2000
temp_2000 <- ggplot(avg_temp_2000) +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "white") +
  geom_point(aes(x = LONGITUDE, y = LATITUDE, color = avg_temp), size = 2) +
  scale_color_viridis_c(option = "plasma", name = "Avg. Temp. (°F)") +
  coord_fixed(1.3) +
  facet_wrap(~season) +
  labs(title = "Seasonal Average Temperatures (2000)",
       x = "",
       y = "",
       tag = "Figure 1: This is a test note") +
  coord_cartesian(clip = "off") +
  theme(plot.title = element_text(hjust = 0.5),
          plot.margin = margin(t = 10, r = 10, b = 40, l = 10),
          plot.tag.position = c(0.2, -0.1)
          ) +
  theme_minimal()

# Create the ggplot for 2024
temp_2024 <- ggplot(avg_temp_2024) +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "white") +
  geom_point(aes(x = LONGITUDE, y = LATITUDE, color = avg_temp), size = 2) +
  scale_color_viridis_c(option = "plasma", name = "Avg. Temp. (°F)") +
  coord_fixed(1.3) +
  facet_wrap(~season) +
  labs(title = "Seasonal Average Temperatures (2024)",
       x = "", y = "") +
  theme_minimal()



# Arrange them to be side by side!


grid.arrange(temp_2000, temp_2024, ncol = 2)

```

b.  Precipitation Difference between 2000 - 2024 based on seasonality

```{r}
#PRECIPITATION!!!

#joining the two to relay to the map
avg_prec_map <- avg_prec %>%
  left_join(station_coords, by = "STATION")

#Creating map of avg temperature for Selected stations based on seasonality!

ggplot(avg_prec_map %>% filter(season_year >= 2000 & season_year <= 2024)) +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "white") +
  geom_point(data = avg_prec_map %>%
               filter(season_year >= 2000 & season_year <= 2024),
             aes(x = LONGITUDE, y = LATITUDE, color = avg_prec), size = 2) +
  scale_color_viridis_c(option = "plasma", name = "Avg. Prec. (100 μm)") +
  coord_fixed(1.3) +
  facet_wrap(~season) +
  labs(title = "Seasonal Average Precipitation by Station (2000–2024)",
       x = "", y = "") +
  theme_minimal()


# Let's compare first just what 2000 vs 2024 looks like!

# separating 2000 and 2024

avg_prec_2000 <- avg_prec_map %>% filter(season_year == 2000)
avg_prec_2024 <- avg_prec_map %>% filter(season_year == 2024)

# Create the ggplot for 2000
prec_2000 <- ggplot(avg_prec_2000) +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "white") +
  geom_point(aes(x = LONGITUDE, y = LATITUDE, color = avg_prec), size = 2) +
  scale_color_viridis_c(option = "plasma", name = "Avg. Prec. (100 μm)") +
  coord_fixed(1.3) +
  facet_wrap(~season) +
  labs(title = "Seasonal Average Precipitation (2000)",
       x = "", y = "") +
  theme_minimal()

# Create the ggplot for 2024
prec_2024 <- ggplot(avg_prec_2024) +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "white") +
  geom_point(aes(x = LONGITUDE, y = LATITUDE, color = avg_prec), size = 2) +
  scale_color_viridis_c(option = "plasma", name = "Avg. Prec. (100 μm)") +
  coord_fixed(1.3) +
  facet_wrap(~season) +
  labs(title = "Seasonal Average Precipitation (2024)",
       x = "", y = "") +
  theme_minimal()



# Arrange them to be side by side!

grid.arrange(prec_2000, prec_2024, ncol = 2)


```

3.  Bird migration pattern time!

Based on eBird data, we are focusing in the Western Pacific region of
the U.S. (Arizona, California, Oregon, and Washington). Based on the
eBird data set variables for understanding migratory codes:

```{r}
# read in bird data:

amebit_data <- read_csv("data/AB Data/amebit_data.csv")
View(amebit_data)
#I already precleaned it because of how much data there was for the American Bittern, so I don't have to clean it that much!


# Cleaning the Data to focus on what we want:

amebit_monthly <- amebit_data %>%
  mutate(ob_date = mdy(`OBSERVATION DATE`),
         year = year(ob_date),
         month = month(ob_date, label = TRUE)) %>%
  group_by(year, month) %>%
  summarize(observations = n(), .groups = "drop") %>%
  filter(year >= 2000, year <= 2024)




# Visualize time:

ggplot(amebit_monthly, aes(x = month, y = observations, group = year, color = as.factor(year))) +
  geom_line(alpha = 0.5) +
  labs(
    title = "Seasonal Observation Trends of American Bittern 2000 - 2024",
    x = "Month",
    y = "Number of Observations",
    color = "Year"
  ) +
  theme_minimal()


```

4.  Looking at seasonal changes over time

```{r}

seasonal_summary <- climate_data %>%
  mutate(
    TAVG_F = (TAVG * 9/5) + 32,       # Convert temp to Fahrenheit
    PRCP_mm = PRCP / 10               # Convert tenths of mm to mm
  ) %>%
  group_by(season_year, season) %>%
  summarize(
    avg_temp = mean(TAVG_F, na.rm = TRUE),
    avg_precp = mean(PRCP_mm, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(season_year >= 2000, season_year <= 2024)


# Reshaping to make plotting easier to understand shifts in Temp and Precip!

long_summary <- seasonal_summary %>%
  pivot_longer(cols = c(avg_temp, avg_precp),
               names_to = "variable", values_to = "value")


# Plotting:

ggplot(long_summary, aes(x = season_year, y = value, color = season)) +
  geom_line(size = 1) +
  facet_wrap(~ variable, scales = "free_y", 
             labeller = as_labeller(c(avg_temp = "Avg Temp (°F)", avg_precp = "Avg Precip (mm)"))) +
  labs(
    title = "Seasonal Trends in Temperature and Precipitation (2000–2024)",
    x = "Year", y = NULL
  ) +
  scale_color_brewer(palette = "Set2", name = "Season") +
  theme_minimal()

```

5.  Let's combine them to see if there is a difference!

```{r}
# Adding seasonality to American Bittern data:

amebit_season <- amebit_monthly %>%
  mutate(season = case_when(
    month %in% c("Dec", "Jan", "Feb") ~ "Winter",
    month %in% c("Mar", "Apr", "May") ~ "Spring",
    month %in% c("Jun", "Jul", "Aug") ~ "Summer",
    month %in% c("Sep", "Oct", "Nov") ~ "Fall"
  )) %>%
  group_by(year, season) %>%
  summarize(total_obs = sum(observations), .groups = "drop")


# Merging the climate data and bittern data

combined_data <- left_join(amebit_season, seasonal_summary, 
                           by = c("year" = "season_year", "season"))


# Visualize temperature and precipitation seasonality and bittern data:

ggplot(combined_data, aes(x = avg_temp, y = total_obs, color = season)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Seasonal Temperature and Bittern Observations",
    x = "Average Temperature (°F)",
    y = "Bittern Observations"
  ) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()

ggplot(combined_data, aes(x = avg_precp, y = total_obs, color = season)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Seasonal Precipitation and Bittern Observations",
    x = "Average Precipitation (100 μm)",
    y = "Bittern Observations"
  ) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()


# Relationship between climate and bittern evolving together!

combined_long <- combined_data %>%
  pivot_longer(cols = c(avg_temp, avg_precp, total_obs),
               names_to = "variable", values_to = "value")

ggplot(combined_long, aes(x = year, y = value, color = season)) +
  geom_line() +
  facet_wrap(~ variable, scales = "free_y",
             labeller = as_labeller(c(
               avg_temp = "Avg Temp (°F)",
               avg_precp = "Avg Precip (mm)",
               total_obs = "Bittern Observations"
             ))) +
  labs(
    title = "Bittern Observations and Climate Trends by Season (2000–2024)",
    x = "Year", y = NULL
  ) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()



```

#### **Results:**

Precipitation in Fall, Winter, and Spring varied greatly from year to
year, and the Fall and Winter seasons show an overall increase in the
past 25 years. Spring and Summer show a less dramatic decrease.

All seasons but Spring showed a several degree increase in average
temperature since 2000,

The correlation graph displays the relationship between average seasonal
temperature (°F) and the number of Bittern observations, with separate
linear trends plotted for each season. The visualized data reveals
distinct seasonal patterns in both temperature and observations.

During Winter, average temperatures ranged from approximately 37°F to
47°F, with Bittern observations generally below 1,000. A slight positive
trend was noted, suggesting a weak relationship between increasing
temperature and higher observations. In Spring, temperatures were around
50°F to 58°F. Although this season had the highest overall number of
Bittern observations—with some counts exceeding 2,000—there was a slight
negative trend in the linear fit, indicating that the number of Bittern
observations tended to decrease slightly with increasing temperature.
Fall showed a moderate positive trend, with observations increasing
alongside temperatures ranging from 52°F to 59°F. However, overall
observations remained lower compared to Spring and Summer. Summer
displayed the strongest positive trend. As temperatures increased from
67°F to 74°F, Bittern observations increased significantly. This season
also showed more consistent and clustered high observations, suggesting
a stronger relationship between higher temperatures and increased
presence of Bitterns.

#### **Discussion:**

These results suggest that seasonal temperature plays a variable role in
Bittern observations, likely reflecting differences in migratory
behavior, breeding activity, and habitat use throughout the year.

The strong positive trend in Summer may indicate that Bitterns are more
active or easier to observe during warmer months, potentially due to
breeding activity or greater availability of wetland habitats.
Similarly, the positive trend in Fall could reflect pre-migration
behavior or favorable foraging conditions as temperatures cool.

The Spring season, despite having the highest individual observation
counts, showed a slight negative trend. This may reflect a peak in
migration or nesting activity occurring at lower spring temperatures,
after which observations decline as temperatures rise.

Winter observations were consistently low, which aligns with the
species' known migratory behavior—many populations leave colder regions
during winter months. The weak positive correlation may suggest that in
relatively milder winters, some individuals remain present or return
earlier.

Overall, this analysis underscores how temperature and seasonality can
influence bird detection and presence. Further investigation into
habitat variables, precipitation, and food availability could clarify
the mechanisms behind these seasonal patterns. Understanding these
relationships is especially important in the context of climate change,
which may shift seasonal temperature regimes and, in turn, influence
Bittern migration and habitat use.

#### **References:**

Status and trends of wetlands in the conterminous United States 2004 to
2009. US Department of the Interior, US Fish and Wildlife Service,
Fisheries and Habitat Conservation.

eBird. (2025). eBird: An online database of bird distribution and
abundance \[web application\]. eBird, Cornell Lab of Ornithology,
Ithaca, New York. Available: http://www.ebird.org. (Accessed: April 16,
2025).

La Sorte, F. A., & Thompson III, F. R. (2007). Poleward shifts in winter
ranges of North American birds. Ecology, 88(7), 1803-1812.
